---
title: "SDS 315- Homework 2"
author: "Raina James (rj24685)"
output: html_document
---
**GitHub Repository**: [SDS-315 Homework 2 Repository](https://github.com/jamesraina/SDS-315--Homework2/blob/main/HW2.Rmd)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
library(dplyr)
library(ggplot2)
library(kableExtra)

```

## Question 1: Analysis of Evaluation Scores

### Part A: Distribution of Course Evaluation Scores

```{r part-a}
data <- read.csv("profs.csv")
eval_mean <- mean(data$eval, na.rm = TRUE)
eval_sd <- sd(data$eval, na.rm = TRUE)

# Plot
ggplot(data, aes(x = eval)) +
  geom_histogram(binwidth = 0.5, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Course Evaluation Scores",
    x = "Evaluation Score",
    y = "Count"
  ) +
  theme_minimal()
```

*This histogram shows the distribution of evaluation scores, with the mean being `r eval_mean`, and the standard deviation being `r eval_sd`. Most of the evaluations cluster between 4 and 5, which shows generally favorable ratings across the courses.*

---

### Part B: Evaluation Scores by Native English-Speaking Status

```{r part-b}
# Summary statistics by native status
data_summary <- data %>%
  group_by(native) %>%
  summarize(mean_eval = mean(eval, na.rm = TRUE), sd_eval = sd(eval, na.rm = TRUE))

# Plot
ggplot(data, aes(x = native, y = eval, fill = native)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  labs(
    title = "Evaluation Scores by Native English-Speaking Status",
    x = "Native English Speaker",
    y = "Evaluation Score"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

*The boxplots show a slightly higher median evaluation scores for native English speakers. The average score for native speakers is `r data_summary$mean_eval[1]`, and for non-native speakers it is `r data_summary$mean_eval[2]`. The overlapping distributions indicate that nativeness has a minimal effect.*

---

### Part C: Evaluation Scores by Gender

```{r part-c}
gender_summary <- data %>%
  group_by(gender) %>%
  summarize(mean_eval = mean(eval, na.rm = TRUE), sd_eval = sd(eval, na.rm = TRUE))

# Plot
ggplot(data, aes(x = eval)) +
  geom_histogram(binwidth = 0.5, fill = "lightblue", color = "black", alpha = 0.7) +
  facet_wrap(~ gender, ncol = 1) +
  labs(
    title = "Evaluation Scores by Gender",
    x = "Evaluation Score",
    y = "Count"
  ) +
  theme_minimal()
```

*Faceted histograms for male and female instructors show overall no significant differences. The average evaluation score for male instructors is `r gender_summary$mean_eval[gender_summary$gender == 'male']` and for female instructors it is `r gender_summary$mean_eval[gender_summary$gender == 'female']`. Both of the genders show clustering around scores of 4 and 5.*

---

### Part D: Attractiveness vs. Evaluation Scores

```{r part-d}
# Correlation
correlation <- cor(data$beauty, data$eval, use = "complete.obs")

# Plot
ggplot(data, aes(x = beauty, y = eval, color = gender)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "black") +
  labs(
    title = "Attractiveness vs. Evaluation Scores",
    x = "Attractiveness (Beauty Score)",
    y = "Evaluation Score"
  ) +
  theme_minimal()
```

*The scatterplot displays a slight positive relationship between attractiveness and evaluation scores. The correlation coefficient is `r correlation`. Though attractiveness may influence higher scores, the impact seem relatively minimal.*

## Question 2: Bike Sharing Analysis

### Plot A: Average Hourly Bike Rentals

```{r plot-a}

bikeshare <- read.csv("bikeshare.csv")
avg_hourly <- bikeshare %>%
  group_by(hr) %>%
  summarize(avg_total = mean(total, na.rm = TRUE))

# Plot
ggplot(avg_hourly, aes(x = hr, y = avg_total)) +
  geom_line(color = "blue", size = 1) +
  labs(
    title = "Average Hourly Bike Rentals",
    x = "Hour of the Day",
    y = "Average Rentals"
  ) +
  theme_minimal()
```

*The line graph displays the average number of bike rentals throughout the day. The x-axis shows the hour of the day (0 to 23) and y-axis shows the average number of rentals. The graph displays peaks around 8 AM and 5 PM, which aligns with typical morning and evening commute hours. This pattern shows that bike-sharing systems play an important role in supporting daily commutes.*

---

### Plot B: Hourly Rentals by Working Day

```{r plot-b}
avg_workingday <- bikeshare %>%
  group_by(hr, workingday) %>%
  summarize(avg_total = mean(total, na.rm = TRUE))

# Plot
ggplot(avg_workingday, aes(x = hr, y = avg_total, color = factor(workingday))) +
  geom_line(size = 1) +
  facet_wrap(~ workingday, labeller = as_labeller(c(`0` = "Non-Working Day", `1` = "Working Day"))) +
  labs(
    title = "Hourly Bike Rentals by Working Day",
    x = "Hour of the Day",
    y = "Average Rentals",
    color = "Working Day"
  ) +
  theme_minimal()
```

*This faceted line graph compares the hour of day for working days and non-working days in bike rentals. X-axis is the hour of day, ranging from 0 to 23, and the y-axis is an average number of rentals. The panel on top is for rentals on working days; the peak around 8 am and 5 pm reflects the commuting hours. The lower panel shows days off, on which rentals build up gradually and peak in the afternoon. Take-home message: This comparison shows how work schedules affect the bike rental patterns, focusing on higher demand for using bikes to go to and from work on working days.*

---

### Plot C: 9 AM Rentals by Weather Situation

```{r plot-c}
rentals_9am <- bikeshare %>%
  filter(hr == 9) %>%
  group_by(weathersit, workingday) %>%
  summarize(avg_total = mean(total, na.rm = TRUE))

# Plot
ggplot(rentals_9am, aes(x = factor(weathersit), y = avg_total, fill = factor(workingday))) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ workingday, labeller = as_labeller(c(`0` = "Non-Working Day", `1` = "Working Day"))) +
  labs(
    title = "9 AM Rentals by Weather Situation",
    x = "Weather Situation",
    y = "Average Rentals",
    fill = "Working Day"
  ) +
  theme_minimal()
```

*The above bar plot gives the average bike rentals for different instances of weather situations in the morning-9 am. X-axis provides the various conditions of the weather (1- Clear, 2 - Mist, 3- Light Snow/Rain and 4 - severe), whereas y-axis shows the average of rental and panel separates working day vs. non-working day. Clear situation represents maximum number of rentals, where the severe condition drastically reduces its usage. Take home lesson: this plot really drives home how the weather affects bike-sharing demand, particularly in the morning peak.*

## Question 3: Capital Metro UT Ridership Analysis

### Plot A: Average Boardings by Hour, Day of Week, and Month

```{r capmetro-a}
capmetro_UT <- read.csv("capmetro_UT.csv")
capmetro_UT <- mutate(capmetro_UT,
  day_of_week = factor(day_of_week, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
  month = factor(month, levels = c("Sep", "Oct", "Nov"))
)

avg_boardings <- capmetro_UT %>%
  group_by(hour_of_day, day_of_week, month) %>%
  summarize(avg_boarding = mean(boarding, na.rm = TRUE))

# Plot
ggplot(avg_boardings, aes(x = hour_of_day, y = avg_boarding, color = month)) +
  geom_line(size = 1) +
  facet_wrap(~ day_of_week, nrow = 2) +
  labs(
    title = "Average Boardings by Hour, Day of Week, and Month",
    x = "Hour of the Day",
    y = "Average Boardings",
    color = "Month"
  ) +
  theme_minimal()
```

*This faceted line graph shows the average boardings throughout the day, faceted by day of the week with lines colored by month. The x-axis shows the hour of the day while the y-axis represents average boardings. Throughout most weekdays, boardings peak at 8 AM and 5 PM - commuting times. However, Monday boardings in September are lower due to the generally slower start of the semester, while Wednesday through Friday counts tend to be lower in November because of the holiday breaks. This is just the pattern during the school year that correlates with holidays and semester dates.*

---

### Plot B: Boardings vs. Temperature by Weekday and Weekend

```{r capmetro-b}
ggplot(capmetro_UT, aes(x=temperature, y=boarding, color=weekend)) +
  geom_point(alpha=0.5) +
  facet_wrap(~ hour_of_day, ncol=4) +
  scale_color_manual(values=c("red", "green"), labels=c("Weekday", "Weekend")) +
  labs(title="Boardings vs. Temperature Faceted by Hour of Day",
       x="Temperature (°F)",
       y="Boardings",
       color="Day Type") +
  theme_minimal()
```

*This scatter plot shows the relationship between temperature (in degrees Fahrenheit, on the x-axis) and the number of boardings (y-axis) for the Capital Metro UT bus system, based on day type (weekdays and weekends). Each point represents boardings in a 15-minute interval. Points are color-coded to differentiate between two days of the week. The scatter plot reveals that while temperature significantly affects bus ridership—showing reduced boardings at extreme temperatures—this trend is consistent across both weekdays and weekends. This insight suggests that weather conditions impact bus demand uniformly, regardless of the day type.*

## Question 4: Wrangling the Billboard Top 100

### Part A: Top 10 Most Popular Songs Since 1958

```{r billboard-a}
billboard <- read.csv("billboard.csv")

# Data Wrangling
top_songs <- billboard %>%
  group_by(performer, song) %>%
  summarize(count = n(), .groups = 'drop') %>%
  arrange(desc(count)) %>%
  slice_head(n = 10)

# Enhanced Table
top_songs %>%
  kbl(caption = "Top 10 Most Popular Songs on the Billboard Top 100 Since 1958",
      col.names = c("Performer", "Song", "Weeks on Chart")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, italic = TRUE)
```

*Table of the top 10 most popular songs on Billboard Top 100 since 1958. Performer, song title, and total number of weeks each song appeared on the chart. Songs are ranked by count of weeks in descending order.*

---

### Part B: Musical Diversity of the Billboard Top 100 Over Time

```{r billboard-b}
# Filter data for complete years only
filtered_billboard <- billboard %>%
  filter(year != 1958 & year != 2021)

# Data Wrangling for Diversity Measure
diversity <- filtered_billboard %>%
  group_by(year) %>%
  summarize(unique_songs = n_distinct(song), .groups = 'drop')

# Plot
ggplot(diversity, aes(x = year, y = unique_songs)) +
  geom_line(color = "blue", size = 1) +
  labs(
    title = "Musical Diversity on the Billboard Top 100 Over Time",
    x = "Year",
    y = "Number of Unique Songs"
  ) +
  theme_minimal()
```

*This line graph looks at the musical diversity of Billboard Top 100 over time through the number of unique songs featuring in the chart every year. The x-axis shows the yearly values, each missing 1958 and 2021 records due to partial data. While the y-axis shows the counts of unique songs, the tendency reveals a peak of the curves that might highlight the shifting paradigms regarding music consumption, perhaps, through changes in business models and music consumption habits.*

---

### Part C: Artists with the Most "Ten-Week Hits"

```{r billboard-c}
# Identify Ten-Week Hits
ten_week_hits <- billboard %>%
  group_by(performer, song) %>%
  summarize(total_weeks = n(), .groups = 'drop') %>%
  filter(total_weeks >= 10)

# Count Ten-Week Hits by Artist
artist_hits <- ten_week_hits %>%
  group_by(performer) %>%
  summarize(ten_week_hits = n(), .groups = 'drop') %>%
  filter(ten_week_hits >= 30) %>%
  arrange(desc(ten_week_hits))

# Plot
ggplot(artist_hits, aes(x = reorder(performer, -ten_week_hits), y = ten_week_hits)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  coord_flip() +
  labs(
    title = "Artists with the Most Ten-Week Hits",
    x = "Artist",
    y = "Number of Ten-Week Hits"
  ) +
  theme_minimal()
```

*This bar plot shows the 19 artists with the most songs that stayed on the Billboard Top 100 for at least ten weeks. The x-axis lists the artists, while the y-axis is the number of ten-week hits. Bars are ordered by number of hits and the horizontal orientation makes artist names readable.*
